Elemental characterization of personal filter samples from the Household Air Pollution Intervention Network (HAPIN) trial for Source Apportionment

### 2024-03-15 code/data_cleaning_code.qmd
Project Part 3
Created a new QMD file: data_analysis_code, and a Supplementary Material word document. Also, the manuscript file has been updated to include the preliminary results. ReadMe files updated and figures/tables edited.

```{r}
library(here)
library(readxl)
library(tidyverse)
```

@@

```{r}
#Add the `arm` and `fueltype` variables to the `concentration` dataframe
concentration <- concentration %>%
  mutate(arm = hapin_filters$arm, fueltype = hapin_filters$fueltype)

@@

```{r}
#Load file to estimate BC uncertainties
atn <- read_csv(here("data", "raw-data", "ATN_summary_SA.csv"))

@@

```{r}
#Estimate the BC uncertainties
uncertainty <- uncertainty %>% 
  mutate(BC = round((1370 * sqrt((atn$ATN)^2 * 0.000005^2 + 0.000079^2 * 1.16^2)), 4))
```

@@

```{r}
#Add the `arm` and `fueltype` variables to the `uncertainty` dataframe
uncertainty <- uncertainty %>%
  mutate(arm = hapin_filters$arm, fueltype = hapin_filters$fueltype)

>>>>

```{r}
library(here)
library(readxl)
library(tidyverse)
library(naniar)
```

@@

```{r}
#Add the `arm` and `fueltype` variables to the `concentration` dataframe
concentration <- concentration %>%
  mutate(arm = hapin_filters$arm, fueltype = hapin_filters$fueltype)
```

I used the `gg_miss_var()` function to check if there are any missing values from the concentrations file. (For the purpose of the final project and the personal project, any missing values will have to be removed).

```{r}
gg_miss_var(concentration)
```

There is one missing value from the `BC` variable. After doing some research, there is no record that the data information from this filter is available, so the best decision was to drop this value. And then, save the df to an RDS file.

```{r}
#Use na.omit() to remove the observation with missing value
concentration <- na.omit(concentration)

@@

```{r}
#Load file to estimate BC uncertainties
atn <- read_csv(here("data", "raw-data", "ATN_summary.csv"))

@@

```{r}
#First I will remove the observation with `filter_id` 3M53864 from the `uncertainty` df, since this observation does not have a BC concentration available
uncertainty <- uncertainty %>% filter(filter_id != "3M53864") %>% 
  mutate(BC = round((1370 * sqrt((atn$ATN)^2 * 0.000005^2 + 0.000079^2 * 1.16^2)), 4))#Estimate the BC uncertainties
```

@@

```{r}
#First I will remove the observation with `filter_id` 3M53864 from the `hapin_filters` df
hapin_filters <- hapin_filters %>% filter(filter_id != "3M53864")

#Add the `arm` and `fueltype` variables to the `uncertainty` dataframe
uncertainty <- uncertainty %>%
  mutate(arm = hapin_filters$arm, fueltype = hapin_filters$fueltype)

### 2024-03-15 code/data_exploring_code.qmd
Project Part 3
Created a new QMD file: data_analysis_code, and a Supplementary Material word document. Also, the manuscript file has been updated to include the preliminary results. ReadMe files updated and figures/tables edited.

```{r}
#Load packages to use
library(readr)
library(here)
library(vtable)
library(corrplot)
library(tidyverse)
library(RColorBrewer)
library(naniar)
```

@@

I will produce two boxplots, that summarize the concentrations categorized by `arm` and `fueltype`. First I will create a dataframe in longer format using `pivot_longer()`. Then creating the plots using `ggplot()`.

@@

#Creating the boxplot of concentrations by fueltype
options(scipen = 999) #to remove scientific notation for smaller values (i.e. changes from 1*10-3 to 0.001)
ggplot(conc_longer)+
  geom_boxplot(aes(x= species, y= concentration, fill= fueltype))+
  scale_fill_brewer(palette = "Paired")+
  labs(x= "Elemental species", y= "Concentration (ug/m3)")+
  scale_y_log10()+
  guides(fill= guide_legend(title= "Fuel type"))+ #Change legend from fuel_type to Fuel type
  theme_bw(base_size = 15)

#Creating the boxplot of concentrations by arm
options(scipen = 999) #to remove scientific notation for smaller values (i.e. changes from 1*10-3 to 0.001)
ggplot(conc_longer)+
  geom_boxplot(aes(x= species, y= concentration, fill= arm))+
  scale_fill_brewer(palette = "Paired")+
  labs(x= "Elemental species", y= "Concentration (ug/m3)")+
  scale_y_log10()+
  guides(fill= guide_legend(title= "Study Arm"))+ #Change legend from fuel_type to Fuel type
  theme_bw(base_size = 15)
```

@@

>>>>

```{r}
#Load packages to use
library(readr)
library(here)
library(vtable)
library(corrplot)
library(tidyverse)
library(RColorBrewer)
library(naniar)
library(ggpubr)
library(rstatix)
```

@@

I will produce two boxplots, that summarize the concentrations categorized by `arm` and `fueltype`. First I will create a dataframe in longer format using `pivot_longer()`. Then creating the plots using `ggplot()`. In addition, I will add the significance test value to the plots to visualize which chemical species are statistically different among groups.

@@

#conduct the t-test of concentration-fuel_type
stat.test <- conc_longer %>% 
  group_by(species) %>% 
  t_test(concentration ~ fueltype) %>% 
  add_significance() #to add the significance values

#Add the position coordinates
stat.test <- stat.test %>% 
  add_x_position(x= "species", dodge = 0.8) %>% #x position is species, and the length of significance bar is 0.8
  add_y_position(y.trans = function(x){log10(x)}) #add y position, in this case specify using a function that it is in log10 scale

#Creating the boxplot of concentrations by fueltype
options(scipen = 999) #to remove scientific notation for smaller values (i.e. changes from 1*10-3 to 0.001)
ggplot(conc_longer)+
  geom_boxplot(aes(x= species, y= concentration, fill= fueltype))+
  scale_fill_brewer(palette = "Paired")+
  labs(x= "Chemical species", y= "Concentration (ug/m3)")+
  scale_y_log10()+
  guides(fill= guide_legend(title= "Fuel type"))+ #Change legend from fuel_type to Fuel type
  theme_bw(base_size = 15)+
  stat_pvalue_manual(stat.test, label = "p.signif", tip.length = 0.001) #Add the significance bars, use 'p.signif' to add *** or 'ns'

#Creating the boxplot of concentrations by arm
options(scipen = 999) #to remove scientific notation for smaller values (i.e. changes from 1*10-3 to 0.001)
ggplot(conc_longer)+
  geom_boxplot(aes(x= species, y= concentration, fill= arm))+
  scale_fill_brewer(palette = "Paired")+
  labs(x= "Chemical species", y= "Concentration (ug/m3)")+
  scale_y_log10()+
  guides(fill= guide_legend(title= "Study Arm"))+ #Change legend from fuel_type to Fuel type
  theme_bw(base_size = 15)+
  stat_pvalue_manual(stat.test, label = "p.signif", tip.length = 0.001) #Add the significance bars, use 'p.signif' to add *** or 'ns'
```

@@

And now producing summary statistics by type of fuel, using the `sumtable()` function.

```{r}
#Produce summary statistics by type of fuel
#Remove arm and fueltype columns
conc <- concentration %>% select(-arm)

#Using the vtable package, which creates a table
sumtable(conc, group = 'fueltype', group.long = T, add.median = T)
```

### 2024-03-15 products/manuscript/Manuscript.qmd
Project Part 3
Created a new QMD file: data_analysis_code, and a Supplementary Material word document. Also, the manuscript file has been updated to include the preliminary results. ReadMe files updated and figures/tables edited.

---
title: "Elemental characterization of personal filter samples from the HAPIN trial for Source Apportionment"
subtitle: "Project: Part 2"
author: Erick E. Mollinedo
date: '`r format(Sys.Date(), "%B %d, %Y")`'
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
---

@@

# Introduction

## Background

Household air pollution (HAP), due to the combustion of solid fuels for cooking and heating, is one of the most prominent environmental health issues in low- and middle-income countries (LMICs).^1^ HAP has been listed as the second environmental health risk factor worldwide, just after outdoor air pollution.^2^ Exposure to HAP has been linked to more than 2.31 million deaths in 2019 and it is associated with adverse cardiovascular and respiratory health outcomes.^3,4^ Particulate matter with an aerodynamic diameter less than 2.5 micrometers (PM~2.5~) is one of the major air pollutants chemically and physically composed of crustal material, black carbon (BC), organic compounds, metals and others. Toxicity from PM~2.5~ depends, among many factors, on the chemical composition of the particles which heavily depend on the sources of emissions.^5,6^

The identification of point sources of pollution is important for the development of strategies to improve air quality. Source apportionment (SA) is a methodology to reconstruct the impact of emissions from different sources, a method applied to different pollutants such as particulate matter.^7^ The most used SA methods includes receptor models, an example of this is the positive matrix factorization (PMF). One advantage of PMF is that it does not require the source profiles prior to analysis and there is no limit on the number of sources, however, it is required the knowledge of potential source profiles. Another important piece before conducting source apportionment is to correctly process the data (which includes concentrations and uncertainties), and to explore associations to try to identify the potential sources. Just a few studies about SA in LMICs have been published, therefore it is important to characterize chemical composition and potential sources of air pollutants in these settings.^8-10^

The Household Air Pollution Intervention Network (HAPIN) trial was a randomized controlled trial where a liquefied petroleum gas (LPG) stove was delivered as an intervention to half of the pregnant women participants, meanwhile the other half remained cooking using their biomass stove. The intervention was evaluated in terms of reduction of exposures and specific health outcomes by comparing the two study groups (Control and Intervention). Baseline measurements were taken to the participants, and then followed-up for about 18 months, conducting exposure measurements three times during pregnancy and three times after the birth of the child. A description of the HAPIN trial and the methods for exposure sampling can be found elsewhere.^11,12^ A source apportionment pilot study from HAPIN was conducted in 2022. This study was conducted in 64 and 59 personal exposure filter samples from the pilot phase in Guatemala and Rwanda, respectively. Twenty-two chemical species were analyzed in all samples using X-ray fluorescence (XRF), a non-destructive spectrometry analytical method, but only 12 species were detected at both study sites in comparable levels. Based on the chemical species detected, four potential sources of PM~2.5~ were identified in both sites. One study caveat is the small sample size and limited assessment of chemical composition and sources between study arms, so further studies to complement the findings are needed.^13^

## Aims and Hypothesis questions

One of the objectives of this project is to create a clean output that consists in a file with adjusted concentrations for all samples and each elemental species, and another file with the adjusted uncertainties for the same samples and elemental species. These outputs will be used as inputs for the EPA PMF 5.0 software to try to determine the potential sources that contributed to these samples. However, the questions I plan to answer with the data analysis of this project is:

1.  What elements are correlated among each other from these filter samples?
2.  Is there a difference in concentrations if we categorize the samples by type of fuel and by study arm?
3.  What exposures could explain the variations in the concentrations of the most significant elemental species?

# Methods

## Data collection

The main exposure data comes from filter samples of personal exposure measurements in pregnant women participants from the HAPIN trial, in Guatemala. In total, six hundred and forty eight sample filters are available, where eighteen correspond to field blanks. From the HAPIN trial study design, in total, six exposure measurements (visits) were conducted during pregnancy in around 800 participants, but for this scenario, only 648 filter samples were randomly selected for chemical species characterization. Other exposures where collected as categorical variables using standardized questionnaires administered in each visit. These exposures capture some of the sources of exposure such as tobacco smoke, trash burning, use on incense or coil, kerosene and others.

## Concentration estimates

The average 24-hour PM~2.5~ concentrations were calculated using the following equation:

@@

The black carbon 24-hour concentrations were calculated using the following equation:

@@

The concentration of the chemical species (elements) on the filters was determined using X-Ray Fluorescence (XRF), and the uncertainties were estimated also based on the XRF instrument.

## Concentration and Uncertainties data processing

@@

## Statistical analysis

The correlation (r^2^) between chemical species will be determined using either individual paired correlations or by a correlation plot matrix. Additional plots, such as boxplots, to visualize and compare how the concentrations differ by type of fuel and by study arm. These differences will be assessed for statistical significance. After exploring the distribution that best fits the dataste, the most appropriate model will be implemented to explore what exposure variables are mostly associated with the concentrations of the most significant elemental species.

# Preliminary Results

## Description of data

1.  Concentrations:

As seen in the following glimpse from the cleaned `concentration` dataframe, it contains the ID of the 630 sample filters, along with the concentrations of 11 chemical species (Mg, Al, Si, S, K, Ca, Ti, Mn, Fe, Zn, and BC), and to which Arm and Fueltype they belong.

```{r}
concentration <- read_rds(here("data", "processed-data", "concentration.rds"))

glimpse(concentration)
```

2.  Uncertainties:

The cleaned `uncertainty` dataframe contains the uncertainty of the 11 chemical species. It contains the same name of variables as the `concentration` dataframe.

```{r}
uncertainty <- read_rds(here("data", "processed-data", "uncertainty.rds"))

glimpse(uncertainty)
```

3.  Other variables from the HAPIN study:

The cleaned `hapin_samples` dataframe contains information of the PM~2.5~, BC and CO (carbon monoxide) concentrations, besides other important recorded exposures, as seen from the glimpse below. It also contains the filter IDs, the arm and fueltype.

```{r}
hapin_samples <- read_rds(here("data", "processed-data", "hapin_samples.rds"))

glimpse(hapin_samples)
```

## Exploratory data analysis

@@

The concentration of the chemical species are distributed in different ranges depending on their type. For example, the highest concentrations were determined to be for Black carbon (BC), with values from 1 to 100 ug/m^3^. The remaining chemical species have lower concentrations, ranging from less than 0.1 ug/m^3^ to 10 ug/m^3^. Other noticeable patterns is that some of the species were found in higher concentrations for personal samples from participants that used biomass stoves, compared to LPG stoves (@fig-concentrations), such as BC, K, and Ca, and similar patterns were observed when categorized by intervention arm (Supplementary material). The following step will be to determine which of these chemical species have statistically significant differences among both comparison groups.

In terms of correlations, Spearman correlation was estimated between all the chemical species and @fig-correlations, shows the associations highlighting the R^2^ value. It is observed that the highest correlation is observed between Si and Fe, Ti-Fe, and Al-Si. These values could indicate that these species come from the same source, so it will probably be easier to identify using the PMF software after the conclusion of this project.

It is important to remember that the concentrations of the chemical species were recalculated as referenced in the `data_cleaning_code.qmd` file, alongside with the uncertainties. A summary statistics table of the uncertainties can be found in the `data_cleaning_code.qmd` code.

@@

### Other Exposure variables

From the `hapin_samples` dataframe, it was observed that there are some missing values for most of the exposure variables according to the Missing variables plot (@fig-missing). The variable with more missing values (\>30) was the CO concentrations. Given that this variable had the highest number of missing concentrations, it may not be ideal to include it as part of the final analysis. Other missing values are from the recorded exposures such as kerosene, trash burning, stove use, external stoves ('stove_other'), and others. It was also observed that there is one observation without BC concentration. This observation could be dropped out since the PMF analysis after the conclusion of this project requires all the concentrations and uncertainties to be available (No NA/missing values).

The bar plot (@fig-categorical), shows that most of the participants recorded little exposures, with the exception of stove use at home ('stove'), stoves from other homes ('stove_other') and trash burning ('trash'). Having not so many recorded exposures other than the previously mentioned, could make the analysis difficult because of sample size or any other statistical adjustments will have to occur before applying a regression model analysis approach.

@fig-pm shows the distribution of PM~2.5~ concentrations by type of fuel. It is observed that the concentrations do not follow a normal distribution and it seems like, as expected, the LPG stove users samples have lower concentrations than the biomass stove users. The similar trends are observed for the CO and BC concentrations (Supplementary material).

![Plot for missing values for the `hapin_samples` dataframe](../../results/figures/missing-values.jpg){#fig-missing}

![Distribution of responses by the Categorical variables](../../results/figures/cat-exposures.jpg){#fig-categorical}

![PM~2.5~ Concentrations](../../results/figures/pm25.jpg){#fig-pm}

@@

# References

1.  Gordon, S. B. et al. Respiratory risks from household air pollution in low and middle income countries. The Lancet Respiratory Medicine 2, 823–860 (2014).
2.  GBD. Global Burden of Disease Study 2019 (GBD 2019) Results. http://ghdx.healthdata.org/gbd-results-tool (2020).
3.  Murray, C. J. L. et al. Global burden of 87 risk factors in 204 countries and territories, 1990–2019: a systematic analysis for the Global Burden of Disease Study 2019. The Lancet 396, 1223–1249 (2020).
4.  Pope, C. A. & Dockery, D. W. Health Effects of Fine Particulate Air Pollution: Lines that Connect. Journal of the Air & Waste Management Association 56, 709–742 (2006).
5.  Clasen, T. et al. Design and Rationale of the HAPIN Study: A Multicountry Randomized Controlled Trial to Assess the Effect of Liquefied Petroleum Gas Stove and Continuous Fuel Distribution. Environ Health Perspect 128, 047008 (2020).
6.  Mohammed, M. O. A. et al. Potential Toxicological and Cardiopulmonary Effects of PM2.5 Exposure and Related Mortality: Findings of Recent Studies Published during 2003-2013. Biomed Environ Sci.
7.  Viana, M. et al. Source apportionment of particulate matter in Europe: A review of methods and results. Journal of Aerosol Science 39, 827–849 (2008).
8.  Zhou, Z. et al. Chemical Characterization and Source Apportionment of Household Fine Particulate Matter in Rural, Peri-urban, and Urban West Africa. Environ. Sci. Technol. 48, 1343–1351 (2014).
9.  Sharma, S. K. et al. Source Apportionment of PM2.5 in Delhi, India Using PMF Model. Bull Environ Contam Toxicol 97, 286–293 (2016).
10. Martínez-Cinco, M., Santos-Guzmán, J. & Mejía-Velázquez, G. Source apportionment of PM 2.5 for supporting control strategies in the Monterrey Metropolitan Area, Mexico. Journal of the Air & Waste Management Association 66, 631–642 (2016).
11. Clasen, T., Checkley, W., Peel, J. L., Balakrishnan, K., McCracken, J. P., Rosa, G., Thompson, L. M., Barr, D. B., Clark, M. L., Johnson, M. A., Waller, L. A., Jaacks, L. M., Steenland, K., Miranda, J. J., Chang, H. H., Kim, D.-Y., McCollum, E. D., Davila-Roman, V. G., Papageorghiou, A., … HAPIN Investigators. Design and Rationale of the HAPIN Study: A Multicountry Randomized Controlled Trial to Assess the Effect of Liquefied Petroleum Gas Stove and Continuous Fuel Distribution. Environmental Health Perspectives, 128(4), 047008. (2020).
12. Johnson, M. A., Steenland, K., Piedrahita, R., Clark, M. L., Pillarisetti, A., Balakrishnan, K., Peel, J. L., Naeher, L. P., Liao, J., Wilson, D., Sarnat, J., Underhill, L. J., Burrowes, V., McCracken, J. P., Rosa, G., Rosenthal, J., Sambandam, S., de Leon, O., Kirby, M. A., … HAPIN Investigators. Air Pollutant Exposure and Stove Use Assessment Methods for the Household Air Pollution Intervention Network (HAPIN) Trial. Environmental Health Perspectives, 128(4), 047009. (2020).
13. Kearns, K. A. Exposures to household air pollution in women across four diverse countries as part of a multi-country cookstove intervention trial. (2022).

# Supplementary material

![Chemical species concentrations by study arm](../../results/figures/boxplot-arm.jpg)

![Carbon monoxide (CO) Concentrations](../../results/figures/co.jpg)

![Black carbon (BC) Concentrations](../../results/figures/bc.jpg)

>>>>

---
title: "Elemental characterization of personal filter samples from the Household Air Pollution Intervention Network (HAPIN) trial for Source Apportionment"
subtitle: "Project: Part 3"
author: Erick E. Mollinedo
date: '`r format(Sys.Date(), "%B %d, %Y")`'
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/sourceapp-references.bib
csl: ../../assets/diabetologia.csl
---

@@

## Introduction

Household air pollution (HAP), due to the combustion of solid fuels for cooking and heating, is one of the most prominent environmental health issues in low- and middle-income countries (LMICs) [@gordon_respiratory_2014]. HAP has been listed as the second environmental health risk factor worldwide, just after outdoor air pollution. Exposure to HAP has been linked to more than 2.31 million deaths in 2019 and it is associated with adverse cardiovascular and respiratory health outcomes [@murray_global_2020; @pope_health_2006]. Particulate matter with an aerodynamic diameter less than 2.5 micrometers (PM~2.5~) is one of the major air pollutants chemically and physically composed of crustal material, black carbon (BC), organic compounds, metals and others. Toxicity from PM~2.5~ depends, among many factors, on the chemical composition of the particles which heavily depend on the sources of emissions [@clasen_design_2020; @mohammed_potential_nodate].

The identification of point sources of pollution is important for the development of strategies to improve air quality. Source apportionment (SA) is a methodology to reconstruct the impact of emissions from different sources, a method applied to different pollutants such as particulate matter [@viana_source_2008]. The most used SA methods includes receptor models, an example of this is the positive matrix factorization (PMF). One advantage of PMF is that it does not require the source profiles prior to analysis and there is no limit on the number of sources, however, it is required the knowledge of potential source profiles. Said requirement to know the potential sources is associated with the profiles of chemical species, given that the concentration of certain organic and inorganic species alongside their uncertainties can be a good estimate of the point sources [@reff_receptor_2007]. Another important piece before conducting source apportionment is to correctly process the data (which includes concentrations and uncertainties), and to explore associations to try to identify the potential sources. Just a few studies about SA in LMICs have been published, therefore it is important to characterize chemical composition and potential sources of air pollutants in these settings [@zhou_chemical_2014; @sharma_source_2016; @martinez-cinco_source_2016].

The Household Air Pollution Intervention Network (HAPIN) trial was a randomized controlled trial conducted in four different LMICs, where a liquefied petroleum gas (LPG) stove was delivered as an intervention to half of the pregnant women participants, meanwhile the other half remained cooking using their biomass stove. The intervention was evaluated in terms of reduction of exposures and specific health outcomes by comparing the two study groups (Control and Intervention). Baseline measurements were taken to the participants, and then followed-up for about 18 months, conducting exposure measurements three times during pregnancy and three times after the birth of the child. A description of the HAPIN trial and the methods for exposure sampling can be found elsewhere [@clasen_design_2020; @johnson_air_2020]. A source apportionment pilot study from HAPIN was conducted in 2022. This study was conducted in 64 and 59 personal exposure filter samples from the pilot phase in Guatemala and Rwanda, respectively. Twenty-two chemical species were analyzed in all samples using X-ray fluorescence (XRF), a non-destructive spectrometry analytical method, but only 12 species were detected at both study sites in comparable levels. Based on the chemical species detected, four potential sources of PM~2.5~ were identified in both sites. One study caveat is the small sample size and limited assessment of chemical composition and sources between study arms, so further studies to complement the findings are needed [@kearns_exposures_nodate].

The following study aims to produce the clean dataset from the HAPIN trial in Guatemala, needed for PMF analysis and to conduct the exploratory and main data analysis to answer some questions before the PMF modelling: 1.) What chemical species are correlated among each other from these filter samples?, 2.) What are the differences in concentrations of the chemical species if the samples are categorized by type of fuel and by study arm? and 3.) What exposures are associated with the variations in concentrations of the most significant chemical species?

## Methods

### Data collection

The main exposure data comes from filter samples of PM~2.5~ personal exposure measurements in pregnant women participants from the HAPIN trial, in Guatemala. The Enhanced Children's MicroPEM (ECM) was the instrument used for the measurement of PM~2.5~ exposures. In summary, the instrument works as an air pump with a standard flow of 0.3 liters per minute, so the particles get into an inlet and get collected into the filter. In total, six hundred and forty eight sample filters are available, where eighteen correspond to field blanks. From the HAPIN trial study design, in total, six exposure measurements (visits) were conducted during pregnancy in around 800 participants, but for this scenario, only 648 filter samples were randomly selected for chemical species characterization. Other exposures where collected as categorical variables using standardized questionnaires administered in each visit. These exposures capture some of the sources of exposure such as tobacco smoke, trash burning, use on incense or coil, kerosene and others.

### Concentration estimates

The filters were pre-weighed (before sampling) in a Sartorius cubis microbalance, and then post-weighed in the same instrument after sampling. The difference in mass deposition was calculated as the post-pre weigh difference. The average 24-hour PM~2.5~ concentrations were calculated using the following equation:

@@

Black carbon was estimated measuring transmissometry via a Magee Sootscan instrument. The black carbon 24-hour concentrations were calculated using the following equation:

@@

The concentration of 22 chemical species (elements) on the filters were determined using X-Ray Fluorescence (XRF), and the uncertainties were estimated also based on the XRF instrument.

### Concentration and Uncertainties data processing

@@

### Statistical analysis

The correlation (r^2^) between chemical species will be determined using either individual paired correlations or by a correlation plot matrix. Additional plots, such as boxplots, to visualize and compare how the concentrations differ by type of fuel. These differences will be assessed for statistical significance using a Welch two-sample T-test. Multivariate regression models will be computed to analyze which exposures could explain the concentrations of the key chemical species. Given that the concentrations can be approximated to a gamma distribution, the generalized linear model with gamma distribution will be used to compute the models. The root mean square error (RMSE) and R-squared will be the parameters to evaluate the most fit model for each of the chemical species.

## Preliminary Results

Six hundred and twenty-nine filter samples were considered in the final analyses. @fig-stats1 presents a summary of the concentrations from the biomass fueltype samples. Meanwhile @fig-stats2 shows the summary of concentrations from the LPG observations. As a note, all the observations of the baseline measurements were categorized as biomass fuel type, since at this point, the LPG intervention was not yet allocated.

![Summary statistics of the concentrations in the biomass group](../../results/tables/table1.jpeg){#fig-stats1}

![Summary statistics of the concentrations in the LPG group](../../results/tables/table2.jpeg){#fig-stats2}

### Chemical Species correlations

In terms of correlations, Spearman correlation was estimated between all the chemical species and @fig-correlations shows the associations highlighting the R^2^ value. It is observed that the highest correlation is observed between Si and Fe, Ti-Fe, and Al-Si. These values could indicate that these species come from the same sources, so it will probably be easier to identify using the PMF software after the conclusion of this project.

@@

The concentration of the chemical species are distributed in different ranges depending on their type. For example, the highest concentrations were determined to be for Black carbon (BC), with values from 1 to 100 ug/m^3^. The remaining chemical species have lower concentrations, ranging from less than 0.1 ug/m^3^ to 10 ug/m^3^. Other noticeable patterns is that some of the species were found in higher concentrations for personal samples from participants that used biomass stoves, compared to LPG stoves (@fig-concentrations). When conducting the two-sided T-test, it was determined that all of the chemical species, except Fe, Al and Zn were statistically significantly different between type of fuel. The differences in concentrations were also analyzed by study arm, however, given that all baseline measurements belonged to biomass fueltype observations, these results were not presented but can be observed in the Supplementary material.

@@

### Chemical species prediction

Nine of the eleven exposures were evaluated as predictors of the chemical species concentrations. The use of a generator machine (for corn mill grinding) and incence were left out since these were the categorical exposure variables with the less variability in answers. The gamma distributed generalized linear models were chosen given the concentrations of all the species do not follow a normal distribution.

Based on the models, type of fuel was a strong predictor of concentration for eight chemical species: Mg, Mn, BC, Ca, K, S, Si and Ti. Kerosene was another predictor of BC and Mg concentrations alongside type of fuel and being exposed to other sources of smoke (such as traffic emissions). Meanwhile, smoking and other sources of smoke were high predictors of K. On the other hand, stove usage, independently if it is biomass or LPG, was associated with concentration of Fe. However, additional analyses have to be conducted for final model selection.

@@

## References
